---
title: Coursera - Practical Machine Learning Project - Quantified Self Movement Data
  Analysis
author: "Charles Bradbury"
output:
  pdf_document: default
  html_document:
    fig_height: 9
    fig_width: 9
---

## Introduction  
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it.  

In this project, we will use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants to predict the manner in which they did the exercise.  

## Data Preprocessing  
```{r, cache = T}
library(caret)
library(corrplot)
library(e1071)
library(randomForest)
library(rpart)
library(rpart.plot)
```
### Download the Data
```{r, cache = T}
trainUrl <-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
trainFile <- "./data/pml-training.csv"
testFile  <- "./data/pml-testing.csv"
if (!file.exists("./data")) {
  dir.create("./data")
}
if (!file.exists(trainFile)) {
  download.file(trainUrl, destfile=trainFile, method="curl")
}
if (!file.exists(testFile)) {
  download.file(testUrl, destfile=testFile, method="curl")
}
```  
### Read the Data
After downloading the data from the data source, we can read the raw training and test files into two data frames.  
```{r, cache = T}
trainDataSet <- read.csv("./data/pml-training.csv")
testDataSet <- read.csv("./data/pml-testing.csv")
dim(trainDataSet)
dim(testDataSet)
```
The training data set contains 19,622 observations and 160 variables. The testing data set contains 20 observations and 160 variables. The "classe" variable in the training set is the outcome to predict. 

### Clean the data
For this step, we will clean and tidy the data by getting rid of observations with missing values and some meaningless variables.
```{r, cache = T}
sum(complete.cases(trainDataSet))
```
Let's remove columns that contain NA missing values.
```{r, cache = T}
trainDataSet <- trainDataSet[, colSums(is.na(trainDataSet)) == 0] 
testDataSet <- testDataSet[, colSums(is.na(testDataSet)) == 0] 
```  
Let's get rid of some columns that do not pertain to the analysis we are performing so that we can get rid of some clutter.
```{r, cache = T}
classe <- trainDataSet$classe
trainRemove <- grepl("^X|timestamp|window", names(trainDataSet))
trainDataSet <- trainDataSet[, !trainRemove]
trainCleaned <- trainDataSet[, sapply(trainDataSet, is.numeric)]
trainCleaned$classe <- classe
testRemove <- grepl("^X|timestamp|window", names(testDataSet))
testDataSet <- testDataSet[, !testRemove]
testCleaned <- testDataSet[, sapply(testDataSet, is.numeric)]
```

Now, the cleaned training data set contains 19,622 observations and 53 variables. The testing data set contains 20 observations and 53 variables.

### Split the data into training and testing sets
Let's split the cleaned training set into a training data set (70%) and a validation/testing data set (30%). We will use the validation/tsting data set to conduct cross validation in future steps.   
```{r, cache = T}
set.seed(22519) # For reproducibile purpose
inTrain <- createDataPartition(trainCleaned$classe, p=0.70, list=F)
trainData <- trainCleaned[inTrain, ]
testData <- trainCleaned[-inTrain, ]
```

## Data Modeling
Let's fit a predictive model for activity recognition using **Random Forest** algorithm. This automatically selects important variables and is robust to correlated covariates & outliers in general. We will use **5-fold cross validation** when applying the algorithm. 
```{r, cache = T}
controlRf <- trainControl(method="cv", 5)
modelRf <- train(classe ~ ., data=trainData, method="rf", trControl=controlRf, ntree=250)
modelRf
```
Let's estimate the performance of the model on the validation data set.  
```{r, cache = T}
predictRf <- predict(modelRf, testData)
confusionMatrix(testData$classe, predictRf)
```
```{r, cache = T}
accuracy <- postResample(predictRf, testData$classe)
accuracy
oose <- 1 - as.numeric(confusionMatrix(testData$classe, predictRf)$overall[1])
oose
```
We can see estimated accuracy of the model is 99.30% and the estimated out-of-sample error is 0.70%.

## Predicting for Test Data Set
Apply the model to the original testing data set downloaded from the data source. Let's remove the `problem_id` column first.   
```{r, cache = T}
result <- predict(modelRf, testCleaned[, -length(names(testCleaned))])
result
```  

## Appendix: Figures
1. Correlation Matrix Visualization  
```{r, cache = T}
corrPlot <- cor(trainData[, -length(names(trainData))])
corrplot(corrPlot, method="color")
```
2. Decision Tree Visualization
```{r, cache = T}
treeModel <- rpart(classe ~ ., data=trainData, method="class")
prp(treeModel) # fast plot
```